Using high-resolution variant frequencies to empower clinical genome interpretation

### 2016-11-02 manuscript/SupplementaryMaterial.Rmd
Reducing word count

>>>>

### Analysis tools

R was used for all analyses, as well as to compile the figures, manuscript and supplementary material, making use of the following packages:
knitr/rmarkdown (https://github.com/yihui/knitr)
knitcitations (http://www.carlboettiger.info/2012/05/30/knitcitations.html)
knitauthors (https://github.com/jamesware/knitauthors)
dplyr (https://github.com/hadley/dplyr)
ggplot2 (http://ggplot2.org/)
packrat (https://rstudio.github.io/packrat/)

### 2016-11-02 manuscript/manuscript.Rmd
Reducing word count

Whole exome and whole genome sequencing have been instrumental in identifying causal variants in Mendelian disease patients `r citep("10.1016/j.ajhg.2015.06.009")`. As every individual harbors ~12,000-14,000 predicted protein-altering variants `r citep("10.1038/nature19057")`, distinguishing disease-causing variants from benign bystanders is perhaps the principal challenge in contemporary clinical genetics. A variant's low frequency in, or absence from, reference databases is recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`. The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through the analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency (AF) filtering in Mendelian variant discovery efforts.

@@

It is intuitive that when assessing a variant for a causative role in a dominant Mendelian disease, the frequency of a variant in a reference sample, not selected for the condition, should not exceed the prevalence of the condition `r citep(c("10.1016/j.ajhg.2016.03.024","10.1126/scitranslmed.aad5169"))`. This rule must, however, be refined to account for different inheritance modes, genetic and allelic heterogeneity, and reduced penetrance. In addition, for rare variants, estimation of true population AF is clouded by considerable sampling variance, even in the largest samples currently available. These limitations have encouraged the adoption of very lenient approaches when filtering variants by AF `r citep(c("10.1016/S0140-6736(14)61705-0","10.1038/ng.3304"))`, and recognition that more stringent approaches that account for disease-specific genetic architecture are urgently needed`r citep("10.1016/j.ajhg.2016.03.024")`.

Here we present a statistical framework for assessing whether rare variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts. We demonstrate that AF cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration when compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants. We present pre-computed AF filtering values for all variants in the Exome Aggregation Consortium database, for comparison with user-defined disease-specific thresholds, which are now available through the ExAC data browser and for download, to assist others in applying this framework.

@@

We do not know the true population AF of any variant, having only an observed AF in a finite population sample. Moreover, confidence intervals around this observed frequency are problematic to estimate given our incomplete knowledge of the frequency spectrum of rare variants, which appears to be skewed towards very rare variants. For instance, a variant observed only once in a sample of 10,000 chromosomes is much more likely to have a frequency < 1:10,000 than a frequency >1:10,000. `r citep("10.1038/nature19057")`

To address this, we begin by specifying a maximum _true_ AF value we are willing to consider in the population (using the equation above), from which we can estimate the probability distribution for allele counts in a given sample size. This follows a binomial distribution, and can be satisfactorily approximated with a Poisson distribution (see **Supplementary Methods**). This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given underlying population frequency. For example, a variant with a true population AF of 0.0001 would be expected to occur in a sample of 100,000 alleles ≤15 times with a probability of 0.95.

@@

To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population AF of 4.0x10^-5^ can be observed in a random population sample of a given size. For a 5% error rate we take the 95th percentile of a poisson distribution with λ = expected allele count, which is given by: sample size (chromosomes) &times; expected population AF (**Methods**). For HCM this gives us a maximum tolerated allele count of 9, assuming 50% penetrance (or 5 for fully penetrant alleles), for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes). The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^; Table 1).

@@

All data required to reproduce these analyses is available at <https://github.com/ImperialCardioGenetics/frequencyFilter>. The manuscript was compiled in R, and source code for the analysis, figures and manuscript, are available at the same location. Curated variant interpretations are deposited in ClinVar *Accession & DOI to be added*. ExAC annotations are available at <https://exac.broadinstitute.org>. Our AF calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.

@@

@@

### Figure 1
> Plot of ExAC allele count (all populations) against case allele count for variants classified as VUS, Likely Pathogenic or Pathogenic in 6179 HCM cases. The dotted lines represent the maximum tolerated ExAC allele counts in HCM for 50% (dark blue) and 100% penetrance (light blue). Variants are colour coded according to reported pathogenicity. Where classifications from contributing laboratories were discordant the more conservative classification is plotted. The inset panel shows the full dataset, while the main panel expands the region of primary interest.

> The clinical utility of stringent allele frequency thresholds. (a) The number of predicted protein-altering variants (definition in Methods) per exome as a function of the allele frequency filter applied. A one-tailed 95% confidence interval is used, meaning that variants were removed from consideration if their AC would fall within the top 5% of the Poisson probability distribution for the user's maximum credible AF (x axis). (b) The odds ratio for HCM disease-association against allele frequency. The prevalence of variants in HCM-associated genes (_MYH7_, _MYBPC3_ and other sarcomeric  (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_, analysed collectively)) in 322 HCM cases and 60,706 ExAC controls were compared for a range of allele frequency bins, and an odds ratio computed (see **Methods**). Data for each bin is plotted at the upper allele frequency cutoff. Error bars represent 95% confidence intervals. The probability that a variant is pathogenic is much greater at very low allele frequencies.

> A flow diagram of our approach, applied to a dominant condition, and using ExAC as our reference sample. First, a disease-level maximum credible population allele frequency is calculated, based on disease prevalence, heterogeneity and penetrance. To evaluate a specific variant, we determine whether the observed variant allele count is compatible with disease by comparing this maximum credible population AF against the (pre-calculated) filtering allele frequency for the variant. *while filtering allele frequency has been pre-computed for ExAC variants, the same framework can be readily applied using another reference sample.

### Table 1  
> Details of the most prevalent pathogenic variants in case cohorts for five cardiac conditions. Shown along with the frequency in cases is the estimated population allele frequency (calculated as: case frequency x disease prevalence x 1/2 x 1/variant penetrance) and the observed frequency in the ExAC dataset. ^\*^As penetrance estimates for individual variants are not widely available, we have applied an estimate of 0.5 across these cardiac disorders (see **Supplementary Information**). HCM - hypertrophic cardiomyopathy; DCM - dilated cardiomyopathy; ARVC - arrhythmogenic right ventricular cardiomyopathy; LQTS - long QT syndrome. Case cohorts and prevalence estimates (taken as the highest value reported) were obtained from: HCM`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, DCM`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90","10.1038/nrcardio.2013.105"))`, ARVC`r citep(c("10.1038/gim.2016.90","10.1016/j.ijcard.2005.12.015"))`, LQTS`r citep(c("10.1016/j.hrthm.2009.05.021","10.1016/j.cjca.2012.07.847"))` and Brugada`r citep(c("10.1016/j.hrthm.2009.09.069","10.1016/j.hlc.2015.07.020"))`.

### Table 2
> Maximum credible population frequencies and maximum tolerated ExAC allele counts for variants causative of exemplar inherited cardiac conditions, assuming a penetrance of 0.5 throughout. CPVT - catecholaminergic polymorphic ventricular tachycardia; FH - familial hypercholesterolaemia. Prevalence estimates (taken as the highest value reported) were obtained from: Marfan`r citep("10.1016/S0140-6736(05)67789-6")`, Noonan`r citep("10.1016/S0140-6736(12)61023-X")`, CPVT`r citep("10.1038/ejhg.2013.55")` and classical Ehlers-Danlos`r citep("10.1097/GIM.0b013e3181eed412")`.

>>>>

Whole exome and whole genome sequencing have been instrumental in identifying causal variants in Mendelian disease patients `r citep("10.1016/j.ajhg.2015.06.009")`. As every individual harbors ~12,000-14,000 predicted protein-altering variants `r citep("10.1038/nature19057")`, distinguishing disease-causing variants from benign bystanders is perhaps the principal challenge in contemporary clinical genetics. A variant's low frequency in, or absence from, reference databases is recognised as a necessary, but not sufficient, criterion for variant pathogenicity `r citep(c("10.1038/gim.2015.30","10.1038/nature13127"))`. The recent availability of very large reference databases, such as the Exome Aggregation Consortium (ExAC) `r citep("10.1038/nature19057")` dataset, which has characterised the population allele frequencies of 10 million genomic variants through analysis of exome sequencing data from over 60,000 humans, provides an opportunity to obtain robust frequency estimates even for rare variants, improving the theoretical power for allele frequency (AF) filtering in Mendelian variant discovery efforts.

@@

It is intuitive that when assessing a variant for a causative role in a dominant Mendelian disease, the frequency of a variant in a reference sample, not selected for the condition, should not exceed the prevalence of the condition `r citep(c("10.1016/j.ajhg.2016.03.024","10.1126/scitranslmed.aad5169"))`. This rule must, however, be refined to account for different inheritance modes, genetic and allelic heterogeneity, and reduced penetrance. In addition, for rare variants, estimation of true population AF is clouded by considerable sampling variance, even in the largest samples currently available. These limitations have encouraged the adoption of very lenient AF filtering approaches `r citep(c("10.1016/S0140-6736(14)61705-0","10.1038/ng.3304"))`, and recognition that more stringent approaches that account for disease-specific genetic architecture are urgently needed`r citep("10.1016/j.ajhg.2016.03.024")`.

Here we present a statistical framework for assessing whether variants are sufficiently rare to cause penetrant Mendelian disease, while accounting for both architecture and sampling variance in observed allele counts. We demonstrate that AF cutoffs well below 0.1% are justified for a variety of human disease phenotypes and that such filters can remove an additional two-thirds of variants from consideration compared to traditionally lenient frequency cutoffs, without discarding true pathogenic variants. We present pre-computed AF filtering values for all variants in the ExAC database, for comparison with user-defined disease-specific thresholds, which are available through the ExAC data browser and for download, to assist others in applying this framework.

@@

We do not know the true population AF of any variant, having only an observed AF in a finite population sample. Moreover, confidence intervals around this observed frequency are problematic to estimate given our incomplete knowledge of the frequency spectrum of rare variants, which is skewed towards very rare variants. For instance, a variant observed only once in a sample of 10,000 chromosomes is much more likely to have a frequency < 1:10,000 than a frequency >1:10,000. `r citep("10.1038/nature19057")`

To address this, we begin by specifying a maximum _true_ AF value we are willing to consider in the population (using the equation above), from which we can estimate the probability distribution for allele counts in a given sample size (see **Supplementary Methods**). This allows us to set an upper limit on the number of alleles in a sample that is consistent with a given underlying population frequency. For example, a variant with a true population AF of 0.0001 would be expected to occur in a sample of 100,000 alleles ≤15 times with a probability of 0.95.

@@

To apply this threshold while remaining robust to chance variation in observed allele counts, we ask how many times a variant with population AF of 4.0x10^-5^ can be observed in a random population sample (**Methods**). At a 5% error rate, this yields a maximum tolerated allele count of 9, assuming 50% penetrance (5 for fully penetrant alleles) for variants genotyped in the full ExAC cohort (sample size=121,412 chromosomes). The _MYBPC3_:c.1504C>T variant is observed 3 times in ExAC (freq=2.49x10^-5^;Table 1).

@@

All data required to reproduce the analysis, figures and manuscript (compiled in R) are available at <https://github.com/ImperialCardioGenetics/frequencyFilter>. Curated variant interpretations are deposited in ClinVar *Details to be added*. ExAC annotations are available at <https://exac.broadinstitute.org>. Our AF calculator app is located at <https://jamesware.shinyapps.io/alleleFrequencyApp/>, with source code available at <http://github.com/jamesware/alleleFrequencyApp>.

@@

### Figure 1
Plot of ExAC allele count (all populations) against case allele count for variants classified as VUS, Likely Pathogenic or Pathogenic in 6179 HCM cases. The dotted lines represent the maximum tolerated ExAC allele counts in HCM for 50% (dark blue) and 100% penetrance (light blue). Variants are colour coded according to reported pathogenicity. Where classifications from contributing laboratories were discordant the more conservative classification is plotted. The inset panel shows the full dataset, while the main panel expands the region of primary interest.

The clinical utility of stringent allele frequency thresholds. (a) The number of predicted protein-altering variants (definition in Methods) per exome as a function of the allele frequency filter applied. A one-tailed 95% confidence interval is used, meaning that variants were removed from consideration if their AC would fall within the top 5% of the Poisson probability distribution for the user's maximum credible AF (x axis). (b) The odds ratio for HCM disease-association against allele frequency. The prevalence of variants in HCM-associated genes (_MYH7_, _MYBPC3_ and other sarcomeric  (_TNNT2_, _TNNI3_, _MYL2_, _MYL3_, _TPM1_ and _ACTC1_, analysed collectively)) in 322 HCM cases and 60,706 ExAC controls were compared for a range of allele frequency bins, and an odds ratio computed (see **Methods**). Data for each bin is plotted at the upper allele frequency cutoff. Error bars represent 95% confidence intervals. The probability that a variant is pathogenic is much greater at very low allele frequencies.

A flow diagram of our approach, applied to a dominant condition, and using ExAC as our reference sample. First, a disease-level maximum credible population allele frequency is calculated, based on disease prevalence, heterogeneity and penetrance. To evaluate a specific variant, we determine whether the observed variant allele count is compatible with disease by comparing this maximum credible population AF against the (pre-calculated) filtering allele frequency for the variant. *while filtering allele frequency has been pre-computed for ExAC variants, the same framework can be readily applied using another reference sample.

### Table 1  
Details of the most prevalent pathogenic variants in case cohorts for five cardiac conditions. Shown along with the frequency in cases is the estimated population allele frequency (calculated as: case frequency x disease prevalence x 1/2 x 1/variant penetrance) and the observed frequency in the ExAC dataset. ^\*^As penetrance estimates for individual variants are not widely available, we have applied an estimate of 0.5 across these cardiac disorders (see **Supplementary Information**). HCM - hypertrophic cardiomyopathy; DCM - dilated cardiomyopathy; ARVC - arrhythmogenic right ventricular cardiomyopathy; LQTS - long QT syndrome. Case cohorts and prevalence estimates (taken as the highest value reported) were obtained from: HCM`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90"))`, DCM`r citep(c("10.1038/gim.2014.205","10.1038/gim.2016.90","10.1038/nrcardio.2013.105"))`, ARVC`r citep(c("10.1038/gim.2016.90","10.1016/j.ijcard.2005.12.015"))`, LQTS`r citep(c("10.1016/j.hrthm.2009.05.021","10.1016/j.cjca.2012.07.847"))` and Brugada`r citep(c("10.1016/j.hrthm.2009.09.069","10.1016/j.hlc.2015.07.020"))`.

### Table 2
Maximum credible population frequencies and maximum tolerated ExAC allele counts for variants causative of exemplar inherited cardiac conditions, assuming a penetrance of 0.5 throughout. CPVT - catecholaminergic polymorphic ventricular tachycardia; FH - familial hypercholesterolaemia. Prevalence estimates (taken as the highest value reported) were obtained from: Marfan`r citep("10.1016/S0140-6736(05)67789-6")`, Noonan`r citep("10.1016/S0140-6736(12)61023-X")`, CPVT`r citep("10.1038/ejhg.2013.55")` and classical Ehlers-Danlos`r citep("10.1097/GIM.0b013e3181eed412")`.

@@
